---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyods

<!-- badges: start -->
[![Project Status: Concept – Minimal or no implementation has been done yet, or the repository is only intended to be a limited example, demo, or proof-of-concept.](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
<!-- badges: end -->

`{tidyods}` is an R package to import cells from an OpenDocument Spreadsheet 
(ODS) file, and provide information about the cell's postion, value types and 
formulas, and provide methods to "rectify" cells back to a 2-dimensional 
data.frame (specifically as a tidyverse tibble).

## Installation

You can install the development version of `{tidyods}` like so:

``` r
remotes::install_github("mattkerlogue/tidyods")
```

## Usage
The `{tidyods}` package exports the following functions for users:

- `read_ods_cells(path, sheet, quick = FALSE)`: extract the cells from a sheet 
  in an ODS file
- `read_ods_sheet(path, sheet, rectify = "simple", base_values = TRUE, quick = FALSE)`: 
  extract cells from an ODS sheet and return as a rectangular dataset, like a 
  spreadsheet
- `ods_sheets(path)`: list the sheets in an ODS fie
- `simple_rectify(path)`: "rectify" a set of cells into a rectangular dataset, 
  like a spreadsheet

The package includes an example ODS with two sheets: `penguins` and `types`.

```{r}
library(tidyods)

example_file <- system.file("extdata", "basic_example.ods", package = "tidyods")

ods_sheets(example_file)
```

The `penguins` sheet is a simple 6 rows by 4 columns sheet that stores the 
output of the following code:

```{r}
palmerpenguins::penguins |>
  tidyr::drop_na() |>
  dplyr::group_by(species, female = sex == "female") |>
  dplyr::summarise(
    dplyr::across(c(bill_length_mm, body_mass_g), mean, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
penguin_sheet <- read_ods_sheet(example_file, "penguins", quick = TRUE)

penguin_sheet
```

While the `types` sheet shows examples of the different ODS data types:
```{r}
example_cells <- read_ods_cells(example_file, "types")

example_cells |> 
  dplyr::filter(row > 1) |>
  dplyr::group_by(col) |>
  dplyr::glimpse()

example_cells |>
  dplyr::filter(row > 1) |>
  dplyr::group_by(col) |>
  dplyr::slice_head(n = 2) |>
  dplyr::select(-cell_type)
```

## Performance
An ODS file is a zipped collection of XML files and associated files. 
`{tidyods}`, like the `{readODS}` package, uses the [xml2](https://xml2.r-lib.org) 
package to process this file. There are three likely sources of performance 
issues: downloading of remote files, unzipping the ODS file, processing rows.

With default settings, performance of `{tidyods}` functions is slower than 
`{readODS}`, however `{tidyods}` provides console messages and progress bars to
the user.

Performance can be improved by setting `quick = FALSE` in `read_ods_cells()` 
and `read_ods_sheet()`, this will extract only a simple text representation of 
the cell, i.e. it will not provide information on cell or value types, formulas 
or the underlying base value for numbers, dates or times. This method also 
ignores replicated white space characters.

To test performance we will use an ODS file published by the UK Government on the
[number of civil servants by postcode](https://www.gov.uk/government/statistics/number-of-civil-servants-by-postcode-department-responsibility-level-and-leaving-cause-2021) (5,544 rows by 11 columns). Analysis of read 
times indicates that using the `quick = TRUE` setting for `read_ods_cells()` 
results in a read that is 10 times faster than the full cell extraction method 
and around 1.5 times faster than reading the sheet with `{readODS}`.

```r
microbenchmark::microbenchmark(
  "tidyods_slow" = tidyods::read_ods_cells(test_ods, "Staff_in_post", 
    quick = FALSE),
  "tidyods_quick" = tidyods::read_ods_cells(test_ods, "Staff_in_post", 
    quick = TRUE),
  "readODS" = readODS::read_ods(test_ods, "Staff_in_post"),
  times = 3
)

#> Unit: seconds
#>          expr     min      lq    mean  median      uq     max neval
#>  tidyods_slow 35.4833 35.7546 35.8626 36.0259 36.0523 36.0786     3
#> tidyods_quick  3.2470  3.2584  3.2935  3.2697  3.3167  3.3638     3
#>       readods  5.1141  5.1606  5.2050  5.2072  5.2505  5.2939     3
```

A cursory test of a huge file that is known to be slow/non-performative with 
{readODS} shows promise.

Performance issues are tracked and discussed in 
[issue #3](https://github.com/mattkerlogue/tidyods/issues/3) 
in the package's GitHub repository.

However, if performance is a major determinant of your workflow (i.e. you have
very large files), its likely more sensible to open the file in a spreadsheet
application, save the file as a `.xslx` format file and read using `{tidyxl}`.

## Related projects
The `{tidyODS}` package is heavily inspired by two existing packages:

- The [`{readODS}`](https://github.com/chainsawriot/readODS) package provides
  functionality to read and write ODS files with R, with the resulting 
  data.frame reflecting the way the data is viewed in a spreadsheet editor 
  application (i.e a two-dimension table structure).
- The [`{tidyxl}`](https://nacnudus.github.io/tidyxl/) package reads cells from 
  Excel files and provides a data.frame where each cell is its own row and 
  columns provide information about the cell's location, value and value type, 
  formula and formatting.
- The [`{unpivotr}`](https://nacnudus.github.io/unpivotr/) package works with 
  datasets of cells (such as those created by tidyxl).

## Philosophy
The `{readODS}` package is the only package on CRAN that handles the reading of 
ODS files, whereas there are more than 20 packages (as at 12 June 2022) on CRAN 
that work with Excel files (of which `{tidyxl}` is one). In some respects this 
is due to the wider usage of Excel files by businesses, governments, academia 
and other organisations to publish and share data. Various governments and
international organisations are starting to mandate the use of OpenDocument 
Format files for publishing of their information.

The initial purpose of `{tidyods}` was to provide a second package to the R 
ecosystem for reading ODS files, in part prompted when encountering an error 
with `{readODS}` and discovering no alternative pacakge was available. For 
example if you run into an error when using the `{readxl}` package you could 
use easily try the`{openxlsx}` or `{xlsx}` instead.

The conceptual code development lead to the creation of an output dataset 
similar to that produced by the `{tidyxl}` package.

Thus, `{tidyods}` extracts cells from an ODS document in the same manner as
`{tidyxl}`, but includes functions to "rectify" the cells so as to also provide
an output similar to that of `{readODS}` (and many of those that read Excel 
files). This is similar to the 
[rectify](https://nacnudus.github.io/unpivotr/reference/rectify.html) function 
in the `{unpivotr}` package.
